Concurrent Programming with CXL

### 2020-06-15 book/paper.tex
Latest

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point.  CXL does not support I/O;
\item Python is object-oriented; CXL is not;
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Try it out:

@@

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  Addresses are lexicographically
ordered accordingly;
\item a context consists of a name tag, the value of the program counter,
the value of the register, and the contents of the stack.  Currently,
contexts are ordered by hash.
\end{itemize}

@@

To iterate through a dictionary, list, or tuple, use:
\texttt{for v in keys(e):~S}.

@@

Clarifications:
\begin{itemize}
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\end{itemize}

>>>>

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point, iterators, or I/O;
\item Python is object-oriented; CXL is not;
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Try it out (here \texttt{\$}
represents a shell prompt):

@@

\chapter{Non-Blocking Synchronization}

\chapter{Barrier Synchronization}

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  Addresses are lexicographically
ordered accordingly;
\item contexts (Appendix~\ref{app:context} are ordered by hash.
\end{itemize}

@@

CXL does not support Python-style iterators.
To iterate through a dictionary, list, or tuple, use:
\texttt{for v in keys(e):~S}.

@@

Clarifications:
\begin{itemize}
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\item The \texttt{Frame} instruction pushes the value of the process
register (\emph{i.e.}, the values of the process variables) onto the
stack.  It initializes the \texttt{result} variable to the empty dictionary.
The \texttt{Return} instruction restores the process register by popping
its value of the stack.
\item All method calls have exactly one argument, although it sometimes
appears otherwise:
\begin{itemize}
\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
\item \texttt{m(a)} invokes method \texttt{m} with argument \texttt{a};
\item \texttt{m(a, b, c)} invokes method \texttt{m} with tuple \texttt{(a, b, c)} as argument.
\end{itemize}
The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
\end{itemize}

\chapter{Contexts and Processes}
\label{app:context}

A context captures the state of a process.  Each time the process
executes an instruction, it goes from one context to another.
All instructions update the program counter, and so no instruction
leaves the context the same.
There may be multiple processes
with the same state at the same time.

A context consists of the following:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
name tag & a dictionary with atoms \texttt{.name} and \texttt{.tag} \\
\hline
program counter & a Program Counter value pointing into the code \\
\hline
end & a Program Counter value indicating the end of the process code \\
\hline
atomic & if non-zero, the process is in atomic mode \\
\hline
stack & a list of CXL values \\
\hline
register & a CXL dictionary mapping atoms (representing process variables) to CXL values \\
\hline
stopped & a Boolean indicating if the context is stopped \\
\hline
\end{tabular}
\vspace{1em}

Details:
\begin{itemize}
\item The name in a name tag is the name of the method that the process
is executing;
\item The tag in a name tag is its argument by default.  Optionally
\texttt{spawn} allows the tag to be specified explicitly;
\item A process terminates when it reaches the \texttt{Return} instruction
of the method or when it hits a fault.  Faults include divide by zero,
reading a non-existent key in a dictionary, accessing a non-existent
variable, as well as when an assertion fails;
\item The execution of a process in atomic mode does not get interleaved
with that of other processes.
\item The register of a process always contains a dictionary, mapping
atoms to arbitrary values.  The atoms correspond to the variable names
in a CXL program.
\end{itemize}

A context is considered \emph{blocked} if there does not exist a
sequence instructions that the process can execute that either writes
to a shared variable (even if that leaves the shared variable unchanged)
or terminates the process.

\chapter{The CXL Machine}

The CXL machine has the following state:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
code & a list of CXL machine instructions \\
\hline
labels & a dictionary of atoms to program counters \\
\hline
variables & a dictionnary mapping atoms to values \\
\hline
ctxbag & a bag of non-stopped contexts \\
\hline
stopbag & a bag of stopped contexts \\
\hline
choosing & if not \texttt{None}, indicates a context that is choosing \\
\hline
failure & Boolean indicating some context has failed \\
\hline
initializing & Boolean indicating that the state is not fully initialized. \\
\hline
\end{tabular}
\vspace{1em}

There is initially a single context with nametag
\texttt{\_\_init\_\_/()} and program counter 0.  It starts executing
in atomic mode until it finishes executing the last instruction in
the code.  All states until then are \texttt{initializing} states.
Other processes, created through \texttt{spawn} statements, do not
start executing until then.

A \emph{micro step} is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

A \emph{macro step} can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Choose}.  Also, any instruction after
a \texttt{Stop} instruction starts a macro step.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.

Executing a CXL program results in a graph where the nodes are CXL
states and the edges are macro steps.  A terminating state is a
state with an empty context bag or a state where all contexts are
blocked.
